Algebraic Filtering of Surfaces from 3D Medical Images with Julia

### 2020-04-25 src/paper.tex
Updates from Overleaf

\documentclass[11pt, oneside]{amsart}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ...
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{hyperref}

%SetFonts
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

@@

% rules for for inputed tables
% the
\usepackage{siunitx}
\usepackage{booktabs}

@@

%%%%%%% M. Jirik added - code highlithing
% \usepackage{minted}
% \usepackage[cache=false]{minted}
\usepackage{listings}
\usepackage[dvipsnames]{xcolor}
% \usepackage{minted}
%SetFonts
\lstdefinelanguage{Julia}%
  {morekeywords={abstract,break,case,catch,const,continue,do,else,elseif,%
      end,export,false,for,function,immutable,import,importall,if,in,%
      macro,module,otherwise,quote,return,switch,true,try,type,typealias,%
      using,while},%
   sensitive=true,%
   alsoother={$},%
   morecomment=[l]\#,%
   morecomment=[n]{\#=}{=\#},%
   morestring=[s]{"}{"},%
   morestring=[m]{'}{'},%
}[keywords,comments,strings]%

\lstset{%
    language         = Julia,
    basicstyle       = \ttfamily,
    keywordstyle     = \bfseries\color{blue},
    stringstyle      = \color{magenta},
    commentstyle     = \color{ForestGreen},
    % commentstyle     = \color{green},
    showstringspaces = false,
}
%%%%%%

\title{Algebraic filtering of surfaces from medical images with Julia}
\author{Miroslav Jirik and Alberto Paoluzzi}
\date{}							% Activate to display a given date or no date

\usepackage{draftwatermark}
\SetWatermarkText{Draft}
\SetWatermarkScale{2.0}
\SetWatermarkColor[rgb]{0.9,0.9,0.9}

@@

\begin{abstract}
In this paper we introduce a novel algebraic \textsc{lar-surf} filter, well founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from  segmentation of a 3D medical image. The input is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse binary vector. The output is produced by a linear mapping between spaces of 3- and 2-chains through the boundary operator $\partial_3:C_3 \to C_2$.  In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non connected set of closed surfaces, i.e.~a set of 2-cycles---using the language of algebraic topology. The only data structures used by this approach are sparse arrays with one or two indices, i.e.~sparse vectors and matrices. This work is based on LAR (Linear Algebraic Representation) methods, and is implemented in Julia language, natively supporting parallel computing on hybrid architectures.
\end{abstract}

\tableofcontents

%\section{}
%\subsection{}

%\input{intro}
\section{Introduction}\label{sec:intro}

Isosurface extraction to produce geometric models of surfaces from volumetric data is important in many aplications. It is often used for indirect visualization of the medical data or for flow modeling \cite{Rohan2018a}. 
In this paper we discuss an approach based on basic algebraic topology and linear algebra, using linear spaces $C_p$ of chains (of cells) of dimension $0 \leq p \leq 3$ and the boundary matrix $[\partial_3] : C_3 \to C_2$.
A decomposition of the 3D image into small submatrices called ``bricks'' is performed, the binary coordinate vector of each interesting chain of voxels is generated, and its boundary is computed by matrix multiplication times the boundary matrix. An embarrassing parallel approach is used to compute the boundary surface patches within each brick, that are finally joined.

The most popular algorithm used for surface extraction is probably Marching Cubes (MC). The algorithm was described by Lorentsen and Cline \cite{Lorensen1987} in 1987. Survey of Marching Cubes algorthm has been published in 2006 \cite{Newman2006}. The algorithm is based on considering the cube defining volume. Each corner vertex of the cube is related to input volumetric data. MC traverse the data and constructs the surface by using lookup table of different triangular faces depending on different patterns of the cube.  Main disadvantages of this method are time requirements, ambiquity and holes generation. Some of them were discovered shortly after the algorithm was introduced. 
Marching Cubes. In 1991 Nielson and Hamman described an Asymptotic Decider to solve the ambiguity problem on the faces of the cube.  Natarajan noted that the ambiguity problem also occurs with uniform samples \cite{Natarajan1994}. In 1995 Chernyaev extended the number of cases to 33 \cite{chernyaev1995marching}. More recently the algorithm was updated by Custodio, Pesco and Silva to enhance the quality of iso-surface triangulation \cite{Custodio2019}.

Some alternative methods have been developed, including a method for surface extraction using particle attraction; a system was described by Crossno and Angel in \cite{Crossno2002}. A graph processing that tracks the boundary cell-face adjacencies is described in \cite{Lachaud2000}. Some parallel algorithms for iso-surface extraction are discussed in \cite{Bajaj2004}.
A completely data-parallel algorithm, implementation written in OpenCL that runs entirely on the GPU is presented in~\cite{Smistad:12}. 
A Linear Algebraic Representation algebraic approach, parallelized using the OpenCL framework on Linux was discussed in~\cite{paodcvjcadanda2015}.

The present paper is organized as follows.
Section~\ref{sec:background} provides the basic topological and geometrical concepts needed to understand the \textsc{lar-surf} method, including the building of boundary matrices, the map from Cartesian indices to linear indices, and the Taubin smoothing method.
Section~\ref{sec:filter} discusses the parametric design of the unit block filtered by the parallel algorithm, including the block decomposition, the sparsity rate of the used sparse arrays, the block-level parallelism.
Section~\ref{sec:julia} is related to the algorithm implementation in Julia, and in particular to a discussion of the parallel workflow.
Section~\ref{sec:examples} presents some examples of algorithm execution on the liver and the hepatic portal system.
Section~\ref{sec:conclusion} shortly describes the next extensions of this approach, in particular to implementation with Julia support for GPU parallelism and the multi-segmentation of Medical images.

%\input{background}
\section{Background}\label{sec:background}

Some basic concepts of solid modeling, and in particular the foundational idea of representation scheme, as well as few basic concepts of algebraic topology, are shortly introduced in this Section, including the computation of the matrix of a boundary operator between chain spaces.

\subsection{Representation Schemes}\label{sec:schemes}

A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, like generated by a formal grammar.
Solid pointsets (i.e., `$r$-sets') are defined~\cite{Requicha:1980:RRS:356827.356833} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space. A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined \emph{cellular complex}. In particular, a boundary representation provides a cellular decomposition of the object's boundary into \emph{cells} of dimension zero (vertices), one (edges), and two (faces). Medical imaging can be classified as \emph{enumerative representation} of cellular decompositions of organs and tissues of interest, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.

\subsection{Linear Algebraic Representation}\label{sec:lar}

The \emph{Linear Algebraic Representation} (\textsc{lar}), introduced in~\cite{Dicarlo:2014:TNL:2543138.2543294}, aims to represent the \emph{chain complex}~\cite{TSAS} generated by a piecewise-linear \emph{geometric complex} embedded either in 2D or in 3D. In few words, it gets a minimal characterization of geometry and topology of a cellular complex, i.e.~the embedding mapping $\mu : C_0 \to \E^d$ of 0-cells (vertices), as well a description of $(d-1)$-cells as subsets of vertices, and is able to return the whole chain complex 
%\[ 
%C_\bullet = (C_p, \partial_p) := 
%C_3 \ 
%\substack{
%\delta_2 \x
%\longleftarrow \[-1mm]
%\longrightarrow \
%\partial_3 
%}
%\ C_2 \ 
%\substack{
%\delta_1 \
%\longleftarrow \[-1mm]
%\longrightarrow \
%\partial_2 
%}
%\ C_1 \ 
%\substack{
%\delta_0 \
%\longleftarrow \[-1mm]
%\longrightarrow \
%\partial_1 
%}
%\ C_0 .
%\]

{   
   \centering\vspace{1mm}
   \includegraphics[width=0.475\textwidth]{figs/complex}

}

\noindent
and, in particular, any basis for linear chain spaces $C_p$, and any linear
boundary/coboundary map \(\partial_p\) and
\(\delta_p=\partial_{p-1}^\top\) between them. The \emph{domain} of \textsc{lar} is the set of \textbf{chain complexes} generated by cell $d$-complexes ($2\leq d\leq 3$). The computer \emph{representations} of \textsc{lar} are \textbf{sparse binary matrices} to represent both the operators and the chain bases. Note that in algebraic topology a $p$-chain is defined as a linear combination of $p$-cells with scalars from a field. When the scalar coefficients are from $\{-1, 0, +1\}$, a chain may represent \emph{any (oriented) subset of cells} from the cellular complex.

We may therefore get the $(p-1)$-boundary $\partial_p c_p$ of \emph{any} $p$-chain $c_p$, by multiplication of the coordinate representation $[\partial_p]$ of the boundary operator times the coordinate representation $[c_p]$ of the chain in terms of such scalars, i.e.~by a  matrix-vector product $ [\partial_p] [c_p] $.

It is possible to show that the \textsc{lar} representation scheme is very expressive, i.e.~that it  has a large domain,  including collections of: line segments, quads, triangles, polygons, meshes;  pixels, voxels, volume images; B-reps, enumerative and decompositive representations of solids. 
In this paper we apply \textsc{lar} methods to computation of boundary representations of solid models from segmentation (labeling) of 3D medical images.
To display a triangulation of  boundary faces  in their proper position in space, the information required is contained in the \emph{geometric chain complex} (GCC):
\[
\mu: C_0\to\E^3,\ (\delta_0, \delta_1, \delta_2)
\qquad\equiv\qquad
\texttt{(geom, top) = (W, (EV, FE, CF))}
\]
The GCC allows to transform the (possibly non connected) boundary 2-cycle of surfaces as a standard B-rep~\cite{shapiroSM:202}. 
The geometry \texttt{geom} is given by
the embedding matrix \texttt{W} of vertices (0-cells), and  topology \texttt{top} by the three sparse matrices (\texttt{EV}, \texttt{FE}, \texttt{CF}) of coboundaries ($\delta_0, \delta_1, \delta_2$) of the chain complex describing a   
space arrangement~\cite{paoluzzi2019finite}.
Note that ordered pairs of letters from \texttt{V,E,F,C}, correspond to \emph{\emph{\texttt{V}}ertices$\to$\emph{\texttt{E}}dges$\to$\emph{\texttt{F}}aces$\to$\emph{\texttt{C}}ells} into the 
\emph{\emph{\texttt{C}}olumn$\to$\emph{\texttt{R}}ow} order of matrix maps of operators.

\subsubsection*{Construction of boundary matrix $\partial_d$}

Once fixed an ordering for all the cells (vertices, edges, pixels, and voxels), i.e.~for 0-, 1-, 2-, and 3-elements of a cell partition \texttt{V,E,F,C} of a 3D image, i.e., once fixed the $p$-bases for linear spaces $C_p$ of $p$-chains $(0\leq p\leq 3)$, we call $M_p = (m_{i,j})$ the \emph{characteristic matrix} of the $p$-basis, expressed as subsets of 0-cells, where  we have that $m_{i,j}=1$ if and only if the $j$-th $0$-cell belongs to the boundary of $i$-th $p$-cell, and $m_{i,j}=0$ otherwise.

Note that the product of binary matrices is not binary, so by computing the (sparse) matrix product $(M_{p-1} M_{p}^t) = (n_{i,j})$, with $n_{i,j} = \sum_{k} m_{i,k}m_{k,j}$, we get for each $n_{ij}$ the \emph{number of vertices} shared by $c_{p-1}^i$ and $c_{p}^j$. When this number equates the cardinality of $c_{p-1}^i$, this elementary chain is contained on the boundary of $c_{p}^j$. In a 3D image, with cubic 3-cells and squared 2-cells in between, everywhere we get $n_{i,j}=4$, we may state $c_{2}^i\subset\partial c_{3}^j$. In this case, by looking in each $j$ column of $M_{2} M_{3}^t = (n_{i,j})$, we have exactly \emph{six rows} where  $n_{i,j} = 4$.

Finally, consider the linear graded boundary operator $\partial_p : C_p \to C_{p-1}$. As such, it contains by columns the representation of domain basis elements, expressed as linear combination of the basis elements of the range space. Therefore, the operator matrix $[\partial_d]$ is readily obtained by setting $[\partial_d](i,j)=1$ iff $n_{i,j}=4$ and $[\partial_d](i,j)=0$ otherwise.  Of course, it will contain six non-zero elements for column.  It may be worth to remember that every 3-cell (voxel) of 3D image has exactly six 2-faces.

It is possible to show that all the interesting relations of incidence/adjacency between cells of different dimensions can be both computed and efficiently queried by pairwise computing some matrix products, with one of terms possibly transposed, using only the two boundary and coboundary operator matrices $[\partial_p]$ and $[\delta_p]$, and where $[\delta_p] = [\partial_p^\top]$. We~may also show that such matrices are \emph{very sparse}, with their sparseness growing rapidly with the dimensions (see Section~\ref{sec:block}). The pattern of non-zeros in matrix $[\partial_3]$ corresponding to a brick of shape $(4,4,4)$ is given in Figure~\ref{fig:boundary_matrix_4x4x4}.

\begin{figure}[tbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.75\linewidth]{figs/boundary_matrix_4x4x4.pdf} 
   \caption{A binary image of the coboundary operator  $\delta_2 = \partial_3^\top : C_2 \to C_3$, built for a small 3D image with shape $(4,4,4)$. Note that the number of rows equates the cardinality $4\times 4\times 4 = 64$ of the voxel set; the number of columns is $d\,n\,(1+n)^{d-1} = 3\times 4\times 25 = 300$. Of course, the number of non-zeros per row (cardinality of the facet set of a single voxel) is six, whereas the number of non-zeros per column is two, but on boundary facets.}
   \label{fig:boundary_matrix_4x4x4}
\end{figure}

\subsection{Multiindices from Cartesian indices}\label{sec:inds-from-cart}

In order to utilize the topological algebra shortly recalled in this paper, we need to explicitly sort the cells of the various dimensions into linearly ordered sequences, possibly according to the linear order their information is linearly accommodated in computer storage.

\subsection{Taubin Smoothing}\label{sec:taubin}

Every boundary chain extracted from an image block $\B(i,j,k,n)$ is a \emph{2-cycle}, i.e., a closed 2-chain---in other words, a 2-chain with empty boundary. Such 2-cycles are joined together by removing the double 2-cells (at the boundaries of adjacent bricks) after having suitably shifted their indices to an unique linear representation of the whole image. The resulting raster surface is made by mutually orthogonal raster facets, that must be smoothed in order to get a fair surface. A linear time and space algorithm for this purpose is the Laplacian smoothing, which iteratively  moves each vertex (0-cell) to the centroid of its neighbors. A well known weakness of this simple algorithm is the asymptotic convergence of the whole mesh to a single point, resulting in unfair size reduction even after few iterations.  Conversely, the Taubin smoothing algorithm~\cite{Taubin:1995:SPA:218380.218473,egst.20001029} alternates two Laplacian smoothing steps with \emph{shrink} and \emph{inflate} effects respectively, with the result of delivering pretty invariant sizes and volume of the smoothed mesh. The best results are obtained on meshes which have small variations of edge length and face angles, like for surfaces extracted from 3D raster images, as in our case.

%\input{design}

\section{Block-parametric design}\label{sec:filter}

\subsection{Block decomposition}\label{sec:block-decomposition}

We assume that medical devices produce 3D images with lateral dimensions that are integer multiples of some powers of two, like 128, 256, 512, etc.
Any cuboidal portion of image is completely determined by the Cartesian indices of its voxels of lowest and highest indices, and extracted by multidimensional array \emph{slicing} as $Image([\ell_x : h_x, \ell_y : h_y, \ell_z : h_z])$.

For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $\B$, called \emph{block}, as a function of its element of the  lowest  block  coordinates $i,j,k\in [1:n]$ and of block dimension $n\in\N$:
\[
\B(i,j,k,n) := Image([in:in+n, jn:jn+n, kn:kn+n]) 
\]

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.5\linewidth]{figs/blocks} 
   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block, of size $n^d=64^2$, is sliced by $\B([2,1,64]) = Image([128:172],[64:128)]$}
   \label{fig:blocks}
\end{figure}

Figure~\ref{fig:blocks} shows the block decomposition in a 2D image, with positive integers $(u,v)$ giving the lateral sizes of image. Note that block sides do not necessarily correspond to image edges.

\subsection{Block operator }\label{sec:block}

\subsubsection*{Chain coordinates }\label{sec:chain-coords}
We are going to treat each image block independently from each other. Hence we map each image block $\B(i,j,k,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n\times n$, using coordinate vectors $c\in \mathbf{2}^{n^d} := \{0,1\}^{n^d}$, where the basis element $c \in C_2$ is mapped via Cartesian-to-linear map to the binary vector 
\[
Image(h,k) \mapsto c_{h,k} := [0 \cdots 0\ 1\ 0 \cdots 0] \in \B^{n\times n}
\]
for each $0\leq h,k \leq n$, and where the (single) unit element is in position $nk + h \leq n\times n$.

Therefore, each pixel (or voxel) in a block image will be seen as a basis binary vector in $C_2$, and each subset of image elements, as the corresponding binary vector in $C_2$, with many ones as the cardinality of the subset.

\subsubsection*{Boundary operator }\label{sec:boundary-operator}
For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.

It is easy to see that the operator's matrix $[\partial_d]$ is \emph{very sparse}, since it contains $2\times d$ non-zero elements (ones) for each column (of length $n^d$), i.e.~4 ones and 6 ones for the 2D and 3D case, respectively. In fact the matrix of a linear operator between linear spaces contains by columns the basis element of the domain space, represented in the target space. In our case, the former is an image element (2-cube or 3-cube), represented as the chain of its boundary---i.e. either a 1-cycle of 4 edges, or  a 2-cycle of 6 faces, respectively.

The number of rows of $[\partial_d]$ equates the dimension of the linear space $C_{d-1}$, i.e.~the number of $(d-1)$-cells---elementary $(d-1)$-chains---in the cellular partition of the image. To compute their number, we act in two steps. (a) First we map one-to-one the $n^d$ $d$-cells with $d$ adjacent $(d-1)$-cells, so getting $d\,n^d$ distinct basis elements of $C_{d-1}$. (b) Then we complete this bases by adjoining $n^{d-1}$ boundary elements for each of the $d$ dimensions of the image, so providing further $d\,n^{d-1}$ basis elements for $C_{d-1}$. The dimension of $C_{d-1}$, and therefore the number of rows of $[\partial_d]$ matrix is $d\,(n^{d-1}+n^{d}) = d\,n\,(1+n)^{d-1}$. The number of column equates the number of basis elements of $C_d$, i.e.~the number $n^d$ of block elements.

\subsubsection*{Sparsity and size of boundary matrix }\label{sec:bm-size}

As we have seen, we have $2d$ non-zero elements for each column of $[\partial_d]$, so that their total number is $2d\,n^d$. The number of matrix element is $d\,n\,(1+n)^{d-1} \times n^d$, giving a ratio of 
\[
\frac{\mbox{non-zero\ elements}}{\mbox{total\ elements}} = 
\frac{2d\times n^d}{d\,n\,(1+n)^{d-1} \times n^d} =
\frac{2}{n+n^d}
\]
Using sparse matrices in CSC (Compressed Sparse Column) format we get a storage size:
\[
mem([\partial_d]_{n^d}) = 2\times \#\mbox{nzero} + \#\mbox{columns} = 2\times 2d\,n^d + n^d = (4d+1)n^d.
\]
In conclusion, for block size $n=64$, the matrix $[\partial_d]$ requires for 2D images $9\times 64^2=36,864$ memory elements, and for 3D images $13\times 64^3=3,407,872$ memory elements. Counting the bytes for the standard implementation of a sparse binary matrix (1 byte for values and 8 bytes for indices) we get $(18d+8)n^d$ bytes, giving $176$\,KB for 2D and $15.872$\,MB for 3D.

%TODO fix the name
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.4\linewidth]{figs/grid.pdf} 
   \caption{example caption}
   \label{fig:example}
\end{figure}

\subsection{Block boundary mapping}\label{sec:block-mapping}

Here we refer directly to the 3D case.
Let us call \emph{segment} the bulk content $S$ of interest within the input 3D image of size $(u,v,w)$. Our aim is to compute the segment boundary $\partial_3 S$. 
First we set the size $n$ of the block, in order to decompose the input $Image(u,v,w)$ into a fair number of blocks
\[
M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} \simeq \frac{uvw}{n^3}.
\] 
Then, we consider each segment portion $c_{i,j,k} = S\cap \B(i,j,k,n)$ and compute its local coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$. This one is a sparse binary vector of length $n^3$. Then, assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$, with $d=3$. Finally, compute a matrix $\T{\textbf{B}}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
\[
\T{\textbf{B}} = [\partial_3(n)]\, \T{S}.
\]
where the boundary matrix has dimension $n^d \times dn(n+1)^{d-1}$.
Of course, the $\T{\textbf{B}}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S \cap \B(i,j,k,n)$, and the number of matrix rows is the dimension $n^d$ of the linear space $C_2$ of 2-chains.

\subsubsection*{Embedding}
A final computational step is needed, in order to embed the 2-chains in $\E^3$ space and to assemble the whole resulting surface. In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one with the vertices of the extracted surface. The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\textbf{B}}$.  The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u$, $0\leq y\leq v$, and $0\leq z\leq w$ for each vertex of the 2-cell.  The mapping can be implemented using a dictionary, that will store the inverse coordinate transformation used at the beginning, i.e.~the one from linear to Cartesian coords, in order of not duplicating the output vertices.

\subsubsection*{Surface assembling}

All boundary surface subsets $B(i,j,k) = \partial_3 S \cap \B(i,j,k)$, provided by  columns of $\T{B}$, are embedded in the same coordinate space. In formal terms, using the standard terminology of LAR scheme: 
\[
\texttt{Lar}(S) := (\texttt{Geom}(S), \texttt{Top}(S)) = (\texttt{V}, \texttt{CV}),
\]
where, with respect to the \emph{chain complex} $C_3\to C_2\to C_1\to C_0$ induced by the input image $Im$ and segment portion $S_{i,j,k}$, we get
\begin{align}
\texttt{Geom} &:= \mu(C_0) = \texttt{V},
\\
\texttt{Top} &:= C_3(S) = \T{S} \mapsto \texttt{CV}.
\end{align}
and
\begin{align}
\texttt{Lar}(B_{i,j,k}) &:= (\texttt{Geom}(B_{i,j,k}), \texttt{Top}(B_{i,j,k})) = (\texttt{W}, \texttt{FW}),
\\
\texttt{Geom} &:= \mu(C_0(B_{i,j,k})) = \texttt{W} \subset \texttt{V},
\\
\texttt{Top} &:= C_2(B_{i,j,k}) = \T{\textbf{B}}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
\end{align}

A translation transformation applied to each vertex subset $\texttt{W}_{i,j,k}$ with translation  vector $\v{t} = [i,j,k]$ will therefore move it in the final space position, so finally giving
\[
\texttt{Lar}(\textbf{B}) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
\]

\subsection{Block-level parallelism}\label{sec:block-parallelism}

In the computational pipeline introduced in this paper, several steps can be efficiently performed in parallel at image-block level, depending on the embarassingly data parallel nature of the problem. In particular, little effort is needed to separate the problem into a number of parallel tasks $S_{i,j,k}$, using multiarray slicing. The granularity of parallelism, depending on the block size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$ depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture. The whole approach is appropriate  for SIMD hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.

%\input{implementation}

\section{Julia implementation}\label{sec:julia}

The computer code is implemented in Julia language~\cite{BEKS14} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.

\subsection{Parallel workflow}\label{sec:implementation}

\subsubsection*{Workflow setup}\label{sec:workflow}
The functions in this preliminary step include:
\begin{enumerate}

\item input of 3D medical image $\mathcal{I}$ with \emph{shape} $(\ell_1, \ell_2, \ell_3)$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = [1,2,\ldots,\ell_k]$;

\item analysis of resources available in the computational environment, including operating system, type and number of compute nodes (processors, cores, GPUs), number of cores per node, RAM and caches amounts;

\item depending on the above, best decision for \emph{size} of 3D image block (or brick) $\mathcal{B}$. With defaults $\emph{size}=64$,  the \emph{number of bricks} will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$. 
Hence the default number of bricks is $n=256$, for standard images $512\times 512\times 256$;

\item computation of the Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the storage of $[\partial_B]$ (for $\emph{size}$ = 64) requires about 45 MB;

\item creation of either a local or distributed \texttt{channel} to implement a producer/consumer model of parallel/distributed computation, depending on available resources;

\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (Julia workers) using the Julia macro \texttt{@eveywhere}.

\end{enumerate}

\begin{figure}[tbp]
\includegraphics[width=\textwidth]{figs/schema_horizontal.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\caption{Workflow of \textsc{lar-surf} algorithm}
\label{fig:schema}
\end{figure}

With $size=64$, the number of non-zeros within the sparse matrix $[\partial(64^3)]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB. The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.

\subsubsection*{Job enqueuing}\label{sec:job-enq}
Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication. Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete. The single work items of this stage follow:
\begin{enumerate}

\item extraction, from image arrays of the block views, depending on 3 Cartesian indices;

\item transform  each block \emph{from global} $[\ell_1]\times[\ell_2]\times[\ell_3]$ to \emph{local coordinates} $[n]\times[n]\times[n]$;

\item further transform of each \emph{foreground voxel} $\nu\in\mathcal{S}\subseteq\mathcal{I}$ from Cartesian to linear coordinates, using the suitable Julia's library functions.

>>>>

\documentclass[9pt,academicons]{article}

\usepackage{CADA}

%%%%%%%%%%%%%%%% original head %%%%%%%%%%%%%%%%%%

% \documentclass[11pt, oneside]{amsart}   	% use "amsart" instead of "article" for AMSLaTeX format
% \usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
% \geometry{letterpaper}                   		% ... or a4paper or a5paper or ...
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{hyperref}

%SetFonts
% \usepackage[utf8x]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage{lmodern}

@@

% rules for for inputed tables
% the
\usepackage{siunitx}
\usepackage{booktabs} % tables

@@

%%%%%%% M. Jirik added - code highlithing
% \usepackage{minted}
% \usepackage[cache=false]{minted}
\usepackage{listings}
\usepackage[dvipsnames]{xcolor}
% \usepackage{minted}
%SetFonts
\lstdefinelanguage{Julia}%
  {morekeywords={abstract,break,case,catch,const,continue,do,else,elseif,%
      end,export,false,for,function,immutable,import,importall,if,in,%
      macro,module,otherwise,quote,return,switch,true,try,type,typealias,%
      using,while},%
   sensitive=true,%
%   alsoother={$},%
   morecomment=[l]\#,%
   morecomment=[n]{\#=}{=\#},%
   morestring=[s]{"}{"},%
   morestring=[m]{'}{'},%
}[keywords,comments,strings]%

\lstset{%
    language         = Julia,
    basicstyle       = \ttfamily,
    keywordstyle     = \bfseries\color{blue},
    stringstyle      = \color{magenta},
    commentstyle     = \color{ForestGreen},
    % commentstyle     = \color{green},
    showstringspaces = false,
}
%%%%%%
% TODO remove
\usepackage{todonotes}

\title{Algebraic Filtering of Surfaces from 3D Medical Images with Julia}
% \author{Miroslav Jirik and Alberto Paoluzzi}
\date{}							% Activate to display a given date or no date

% \usepackage{draftwatermark}
% \SetWatermarkText{Draft}
% \SetWatermarkScale{2.0}
% \SetWatermarkColor[rgb]{0.9,0.9,0.9}

@@

% ############################################
% Here is the fun part. Please enter your:            
%      Name (First, Middle initial, Last)                  
%      Your ORCID number replace the "000-0000-1234-5678" part.    
% ############################################

\authorSection{
	\anAuthor{Miroslav Jirik}{0000-0002-8002-2079}{1,3},
	\anAuthor{Alberto Paoluzzi}{0000-0002-3958-8089}{2}, 
% 	\anAuthor{First M. Last}{000-0000-1234-5678}{3} 
}

% ######################################
% Here we need your affiliation and contact e-mail. Please edit   
% affiliation as well as the e-mail fields.                                  
% ######################################
\affiliationSection{
	\anAffiliation{1}{
	Charles University,
    Faculty of Medicine in Pilsen, 
    Biomedical Center}{jirik@lfp.cuni.cz}
	\anAffiliation{2}{Roma Tre University, Department of Mathematics and Physics}{paoluzzi@dia.uniroma3.it}
	\anAffiliation{3}{NTIS, Faculty of Applied Sciences, University of West Bohemia}{}
}

% ######################################
% Please decide on who the corresponding author is going to be 
% and complete the section below                                            
% ######################################
\correspondingAuthor{Miroslav Jirik}{jirik@lfp.cuni.cz}

% ########################################
% Please type in your abstract below after the "\abstract{" part.
% ########################################

\abstract{
In this paper we introduce a novel algebraic 
% \textsc{lar-surf} 
filter, well-founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from  the segmentation of a 3D medical image. 
% TODO 3D medical image
The input of our Linear Algebraic Representation Surface extraction filter (\textsc{lar-surf}) is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse binary vector. The output is produced by a linear mapping between spaces of 3- and 2-chains through the boundary operator $\partial_3:C_3 \to C_2$.  
% In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non-connected set of closed surfaces, i.e. quadrilateral faces or~a set of 2-cycles---using the language of algebraic topology. 
The only data structures used by this approach are sparse arrays with one or two indices, i.e.~sparse vectors and matrices. This work is based on Linear Algebraic Representation methods and is implemented in Julia language, natively supporting parallel  computing on hybrid hardware architectures.
}

\keywords{
Medical 3D, Computational Topology, Linear Algebraic Representation, LAR, Julia, Surface Extraction
}

% DOI: 10.14733/cadconfP.2020.xxx-yyy
\doi{10.3722/cadaps.2020.xxx-yyy}

% \tableofcontents

\input{intro}
\input{background}
\input{design}
\input{implementation}
\input{examples}
\input{conclusion}

\section*{ACKNOWLEDGEMENTS}
\sloppy{
This work was supported by Charles University Research Centre program UNCE/MED/006 ``University Center of Clinical and Experimental Liver Surgery'' 
and Ministry of Education project ITI CZ.02.1.01/0.0/0.0/17\_048/0007280: Application of modern technologies in medicine and industry.
} 
The research was also supported by the project LO 1506 of the Czech Ministry of Education, Youth and Sports.
% The authors appreciate the access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum  provided under the program ``Projects of Large Research, Development, and Innovations Infrastructures'' (CESNET LM2015042).

\appendix
\input{appendix}

\section*{ORCID}
\orcid{Miroslav Jirik}{0000-0002-8002-2079}
\orcid{Alberto Paoluzzi}{0000-0002-3958-8089}

% \clearpage{}
\referenceSection
\bibliographystyle{CADA}
% \bibliography{library,TSAS-2019,references}
\bibliography{TSAS-2019,references}

\bigskip
\end{document}
